A supertree pipeline for summarizing phylogenetic and taxonomic information for millions of species

### 2017-01-30 doc/paper1-submission/paper5-adjusted.tex
Remove appendices.

This operation is performed by the \texttt{otc-uncontested-decompose}
tool in otcetera; see appendix \ref{sec:Decompose-algorithm} for
a description of the algorithm. The output is a series of subproblems,
each of which corresponds to a slice of the taxonomy and corresponding
slices through each relevant input tree. Each uncontested non-terminal
and non-root taxon will show up in two subproblems: it will be the
root of its own subproblem and it will be tip in the subproblem that
covers the next slice deeper in the tree. The red edge in Fig. \ref{fig:exemplify-example}c
highlights the taxa that are not contested by the input shown in Fig. \ref{fig:exemplify-example};
Fig. \ref{fig:subproblems-example} shows the subproblems that would
be emitted as a result of this set of inputs. The supertree operation
of \citet{HinchliffEtAl2015} also used this otcetera-based decomposition
step.

@@

\appendix

\section{The maximum sum displayed groups' weighted scores criterion}\label{sec:The-maximum-sum}

Let $\mathcal{W}$ is a weighting function that maps any input tree's
internal node to a non-negative number. If $I(\mathbb{S},i,j)$ is
an indicator function that is 1 if summary tree $\mathbb{S}$ displays
the node $V(i,j)$ and 0 otherwise then: $SDGWS(\mathbb{S})=\sum_{i}\sum_{j}I(\mathbb{S},i,j)\mathcal{W}(i,j)$
is the ``sum of displayed groups' weighted scores'' for a tree where
$i$ indexes all of the input trees and $j$ indexes each non-root
internal node in tree $i$. Preference for this tree is referred to
as the maximum sum displayed groups' weighted scores criterion (MSDGWS
criterion). The summary tree constructed by the propinquity pipeline
is a greedy heuristic for finding a tree that maximizes this score
when the weights for a node are determined by the tree's weight and
the difference in weighting is so large that displaying one node from
a highly ranked tree is preferred to displaying all of the nodes in
the trees with lower rank.

\section{Description of the decomposition algorithm of otc-uncontested-decompose}\label{sec:Decompose-algorithm}

The input is the a ranked list of input trees and comprehensive taxonomy.

\subsection{Creation of multigraph of the taxonomy with embedded trees}

The tool creates a multigraph by starting with a graph isomorphic
to the taxonomic tree. The nodes and edges created in this step will
be referred to as the ``taxonomic graph.'' Next, we add nodes and
edges to that graph in a procedure that we refer to as ``embedding''
the input trees into the taxonomy. A node is introduced for each node
in an input tree, and these nodes are mapped the MRCA nodes in the
taxonomic graph. In other words, for any node $y$ in an input tree
with a cluster of descendants, $\mathcal{C},$ we find the most tipward
node $z$ in the taxonomic graph that is an ancestor to all of the
taxa in $\mathcal{C}$; let $m(y)=z$ refer to this mapping, and $m^{\prime}(z)=y$
refer to the reverse mapping. Each edge $e_{ij}$ in a source tree
$i$ connects ancestor to its descendant, $a(e_{ij})\rightarrow d(e_{ij})$.
The edges are introduced into the graph. We also introduce new edges
to create a from $m(a(e_{ij}))$ through its descendants to $m(d(e_{ij}))$;
we denote this path $p(e_{ij})$ and refer to the edges in the path
as ``embedding edges for tree $i$''. Note, that this is a path
through the taxonomic nodes, while the edge $e_{ij}$ connects source
tree nodes. The mapping between $e_{ij}$ and $p(e_{ij})$ is stored,
and the edges are labelled with the index $i$ so that it is clear
which tree created them. Because the taxonomic tree is highly unresolved,
it is frequently the case that $m(a(e_{ij}))$ is the same node as
$m(d(e_{ij}))$; in these cases the embedding edge is a loop. This
situation occurs whenever edge $e_{ij}$ can resolve part of the polytomy
represented by a taxon.

We treat the taxonomy as the lowest ranked input tree. The next step
will collapse contested edges in the taxonomic graph. To retain all
of the information from the taxonomy we embed the taxonomy into the
taxonomic graph as if it were another input tree

\subsection{Detection of uncontested higher taxa}

After every input tree and the taxonomy tree have been embedded into
the taxonomic graph, we perform postorder traversal over the taxonomic
graph that underlies the multi-graph. For any internal node (each
of which corresponds to a non-terminal taxon) we determine whether
or not it is contested by examining each input tree. We can determine
tree $i$ contests the taxon represented by taxonomic node $x$ by
looking at the parents of all of the ``exiting'' embedding edges
for tree $i$. These are the set of embedding edges that have $x$
as a daughter and have a parent node that is not $x$ (ergo a parent
node that is taxonomically higher than $x$). If there are more than
one parent nodes in this set of exiting embedding edges, then tree
$i$ contests that taxon. If there is only one parent node, then the all
of the constituent taxa belonging to this taxon that are present in
tree $i$ have one parent that is more inclusive; this means that
the input tree does not contest monophyly of the taxon.

If the taxon is uncontested, then it may be the case that some input
trees do not contest the taxon, but contain polytomies that could
be resolved to display the taxon. The cases can be identified by finding
multiple exiting embedding edges that have the same taxon $y$ as
their parent node. In these cases, a pseudo input tree node is created
and becomes the parent node for these edges; this new node is then
connected to $y$ as if it had been an input edge. This operation
is equivalent to resolving an input tree's polytomy in favor of the
monophyly of the uncontested taxon. This is the only way in which
the input trees are modified during the decomposition.

\subsection{Collapsing contested taxa from the taxonomic graph}

If a taxonomic node $x$ fails the ``uncontested'' test described
in the previous section, then the node corresponding to the taxon
is removed from the taxonomic graph and the set of edges (and mappings
between input edges and embedded paths) is updated as if this taxon
had not been present in when the taxonomic graph was created. This
consists of detecting changing any edge in an embedding path that
is adjacent to $x$ by replacing the reference to $x$ with a reference
to its parent $a(x)$. Note that we do not collapse the edge corresponding
to this taxon in the part of the graph that represents the embedding
of the taxonomy into the taxonomic graph. Thus, the taxonomy will
still claim the monophyly of the taxon. This is relevant if the input
grouping that contests taxon $x$ is overruled (in the subproblem
solution step) by a higher ranked split. In other words, the fact
that the a taxon is contested during the decomposition is not a guarantee
that the taxon will not be monophyletic in the final supertree.

\subsection{Emitting subproblems}

Whenever an uncontested taxon is identified, the appropriate slice
of each input trees that intersect with the taxon is written to a
file. Then the multigraph is simplified by slicing any off the taxon.
This slicing is accomplished by examing all of the exiting embedding
edges for the taxon. The descendant taxa of each input tree is relabeled
with the identifier of the contested taxa and all of that node's descendants
are removed. Thus this input node will act as as if it were a leaf
mapped to the uncontested taxon. All descendants of the taxonomic
graph are also pruned off.
\end{document}

>>>>

This operation is performed by the \texttt{otc-uncontested-decompose}
tool in otcetera; see appendix B for
a description of the algorithm. The output is a series of subproblems,
each of which corresponds to a slice of the taxonomy and corresponding
slices through each relevant input tree. Each uncontested non-terminal
and non-root taxon will show up in two subproblems: it will be the
root of its own subproblem and it will be tip in the subproblem that
covers the next slice deeper in the tree. The red edge in Fig. \ref{fig:exemplify-example}c
highlights the taxa that are not contested by the input shown in Fig. \ref{fig:exemplify-example};
Fig. \ref{fig:subproblems-example} shows the subproblems that would
be emitted as a result of this set of inputs. The supertree operation
of \citet{HinchliffEtAl2015} also used this otcetera-based decomposition
step.

@@

\end{document}
